---
title: "Compiling list of medications based on BNF codes for dementia/AD & palliative care drugs"
author: "Jan Savinc"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Introduction

For the analysis of prescription/dispensing data, lists of specific medicines are needed to make the analysis tractable and applicable to dementia/Alzheimer's disease and for the analysis of anticipatory medicines used in palliative care.

## Source of BNF codes

One source for a list of all BNF codes is the Information Services Portal of the NHS Business Services Authority, https://applications.nhsbsa.nhs.uk/infosystems/data/showDataSelector.do?reportId=126. This is available after putting in a captcha-like password, so I've downloaded the file manually.

File used: `01-01-2023 : 84`
  
# Load packages

```{r}
library(tidyverse)
library(readxl)  # for loading excel files
library(janitor)  # for cleaning data
library(knitr)  # for pretty tables in the report
```


# Download file

Note: this was done manually from the above website!

# Load BNF data

```{r}
## the .csv file is inside a .zip file!
bnf <- read_csv(file = unzip( "./bnf/20230104_1672833147341_BNF_Code_Information.zip", "20230104_1672833147341_BNF_Code_Information.csv")) %>%
  clean_names()
```

# Defining medicines

## Helper functions for searching BNF

```{r}
find_substance <- function(keyword) {
  bnf %>%
    filter(str_detect(bnf_chemical_substance, pattern = regex(keyword, ignore_case = TRUE))) %>%
    group_by(bnf_chemical_substance, bnf_paragraph) %>%
    summarise(
      n = n(),
      products = paste(bnf_presentation, collapse = ";")
    )
}
```


## Dementia / Alzheimer's disease

The medicines of interest are:

- donepezil
- galantamine
- rivastigmine
- memantine

~~I can manually search for the above keywords in the `chemical substance` field in the BNF and include all relevant medicines.~~

Note that the above drugs correspond to the entirety of the BNF section `4.11, Drugs for dementia`

```{r}
bnf_dementia <- 
  bnf %>%
  filter(bnf_section == "Drugs for dementia") %>%
  mutate(medication = word(bnf_chemical_substance, 1)) %>%
  select(medication, bnf_section, bnf_section_code, bnf_chemical_substance_code) %>%
  distinct
```


## Palliative care / anticipatory medicines

Marie Curie lists [the following anticipatory medicines](https://www.mariecurie.org.uk/professionals/palliative-care-knowledge-zone/symptom-control/anticipatory-medicines):

- Medicine for pain in palliative care – an appropriate opioid, for example, morphine, diamorphine, oxycodone or alfentanil.
- Medicine for breathlessness – midazolam or an opioid.
- Medicine for anxiety – midazolam.
- Medicine for delirium or agitation – haloperidol, levomepromazine, midazolam or phenobarbital.
- Medicine for nausea and vomiting – cyclizine, metoclopramide, haloperidol or levomepromazine.
- Medicine for noisy chest secretions – hyoscine hydrobromide or glycopyrronium.

Some of the above are specific chemicals, and some are classes of medicines (e.g. opioids). For opioids, some examples are listed, but I will include the entire class, i.e. the BNF section `4.7.2, Opioid analgesics`. For the other medicines, I will only include the specific substances listed.

Note that alfentanil is not listed as an opioid in the BNF but as a general anaesthetic!

```{r}
bnf_palliative <-
  ## pain: opioids + alfentanil
  list(
    bind_rows(
      bnf %>% filter(bnf_paragraph_code == "040702"),  # opioids
      bnf %>% filter(str_detect(bnf_chemical_substance, pattern = regex("alfentanil", ignore_case = TRUE)))
    ) %>% mutate(medicine = "Pain")
    ,
    ## breathlessness: midazolam or opioid
    bind_rows(
      bnf %>% filter(bnf_paragraph_code == "040702"),  # opioids
      bnf %>% filter(str_detect(bnf_chemical_substance, pattern = regex("midazolam", ignore_case = TRUE)))
    ) %>% mutate(medicine = "Breathlessness")
    ,
    ## anxiety
    bnf %>% filter(str_detect(bnf_chemical_substance, pattern = regex("midazolam", ignore_case = TRUE))) %>% 
      mutate(medicine = "Anxiety")
    ,
    ## delirium/agitation
    bnf %>% filter(str_detect(bnf_chemical_substance, pattern = regex("haloperidol|levomepromazine|midazolam|phenobarbital", ignore_case = TRUE))) %>% 
      mutate(medicine = "Delirium or agitation")
    ,
    ## nausea or vomiting
    bnf %>% filter(str_detect(bnf_chemical_substance, pattern = regex("cyclizine|metoclopramide|haloperidol|levomepromazine", ignore_case = TRUE))) %>% 
      mutate(medicine = "Nausea or vomiting")
    ,
    ## noisy chest secretions
    bnf %>% filter(str_detect(bnf_chemical_substance, pattern = regex("hyoscine hydrobromide|glycopyrronium", ignore_case = TRUE))) %>% 
      mutate(medicine = "Noisy chest secretions")
  ) %>%
  bind_rows() %>%
  select(medicine, bnf_chemical_substance, bnf_chemical_substance_code) %>%
  distinct %>%
  pivot_wider(names_from = medicine, values_from = medicine, names_prefix = "bin_", values_fill = FALSE, values_fn = ~TRUE) %>%
  clean_names()
```

# Conclusion

We are now left with two files listing medicines used in dementia/AD and palliative/anticipatory care; the palliative list is slightly more complex and it includes, for each substance, a number of binary variables indicating its use (e.g. pain, breathlessness, agitation, etc.)

# Save .csv files

```{r}
write_csv(bnf_dementia, file = "./bnf/bnf_codes_dementia_alzheimers.csv")

write_csv(bnf_palliative, file = "./bnf/bnf_codes_palliative_anticipatory.csv")
```

