---
title: "Tables & figures for paper reporting increase of home deaths 2020"
author: "Jan Savinc"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Load packages

```{r, warning=FALSE, message=FALSE}
## install packages from github if not yet available!
# remotes::install_github("datasciencescotland/opendatascot", force = TRUE)
# remotes::install_github("Health-SocialCare-Scotland/phsmethods", force = TRUE)

library(tidyverse)  # for tidy workflow
library(opendatascot)  # importing data from ScotGov open data website
library(phsmethods)  # methods for working with PHS data
library(readxl)  # for reading excel files
# library(SPARQL)  # taken care of by opendatascot
library(lubridate)  # dealing with dates
library(janitor)  # for cleaning column names
library(ISOweek)  # for computing date from ISO week number + year
library(sf)  # for mapping
library(ggrepel)  # for 'mapping'repelling' labels/texst in ggplot
library(patchwork)  # for assembling plots
library(extrafont)  # for working with fonts
library(openxlsx)  # for creating xlsx files
library(scico)  # for continuous grayscale palette
```


# Load workspace

This was previously saved!

```{r}
load(file = "./workspace.RData")
```


# Set up output directory

Because data are updated on a weekly basis, we'll have a separate directory for every week the data are updated:

```{r}
dir_paper <- paste0("./outputs/paper_",latest_data_date)
if (!dir.exists(dir_paper)) dir.create(dir_paper)
```


# Tables

Note: non-covid deaths are defined as all-cause deaths minus covid-related deaths. Otherwise the number of non-covid deaths only includes non-covid deaths once covid-related deaths started being reported in 2020 (week 12).

## Number & proportion of deaths by year & place

```{r}
table_of_all_cause_deaths_by_place_of_death <-
  merged_proportions_of_deaths_by_place %>%
  rename(n = number_of_deaths, prop = proportion_of_total) %>%
  pivot_wider(names_from = c(place_of_death), values_from = c(n, prop), names_sort = FALSE) %>%
  arrange(desc(year)) %>%
  relocate(1,2,3,7,4,8,5,9,6,10)  # manually reorder columns

table_of_non_covid_deaths_by_place_of_death <-
  merged_proportions_of_non_covid_deaths_by_place %>%
  rename(n = number_of_deaths, prop = proportion_of_total) %>%
  pivot_wider(names_from = c(place_of_death), values_from = c(n, prop), names_sort = FALSE) %>%
  arrange(desc(year)) %>%
  relocate(1,2,3,7,4,8,5,9,6,10)  # manually reorder columns

table_1_for_paper <-
  bind_rows(
    table_of_all_cause_deaths_by_place_of_death %>% mutate(cause = "All Deaths") %>% relocate(cause),
    table_of_non_covid_deaths_by_place_of_death %>% mutate(cause = "Non-covid Deaths") %>% relocate(cause)
  )

write.xlsx(x = table_1_for_paper, file = paste0(dir_paper,"/number_of_deaths_by_year_place_pre_edit.xlsx"))
```


## Correlation between annual covid cases & annual deaths

The interesting correlation is between the ratio of 2020 home deaths to historical mean, and the covid cases, rather than the total home deaths!

```{r}
home_deaths_and_cases_2020_hb <-
  total_deaths_hb %>%
  filter(str_starts(place_of_death, "Home") & year == 2020) %>%
  left_join(
    weekly_covid_cases_hb %>%
      filter(year == 2020 & week_number == 53),
    by = c("year", "ref_area")
  ) %>%
  mutate(deaths_non_covid_2020 = deaths_all_causes - deaths_covid_related)

home_deaths_and_cases_2020_la <-
  total_deaths_la %>%
  filter(str_starts(place_of_death, "Home") & year == 2020) %>%
  left_join(
    weekly_covid_cases_la %>%
      filter(year == 2020 & week_number == 53),
    by = c("year", "ref_area")
  ) %>%
  mutate(deaths_non_covid_2020 = deaths_all_causes - deaths_covid_related)

# Non-covid home deaths not included because nearly identical to all deaths


correlations_home_deaths_and_cases <-
  bind_rows(
    broom::tidy(cor.test(home_deaths_and_cases_2020_hb$ratio_annual_deaths_to_historical_mean, home_deaths_and_cases_2020_hb$cumulative_positive, method = "pearson")) %>% mutate(x="Ratio home deaths 2020 to historical mean", y = "Total cases", region = "HB"),
    broom::tidy(cor.test(home_deaths_and_cases_2020_hb$ratio_annual_deaths_to_historical_mean, home_deaths_and_cases_2020_hb$crude_rate_positive_per100k, method = "pearson")) %>% mutate(x="Ratio home deaths 2020 to historical mean", y = "Cases per 100k", region = "HB"),
    broom::tidy(cor.test(home_deaths_and_cases_2020_la$ratio_annual_deaths_to_historical_mean, home_deaths_and_cases_2020_la$cumulative_positive, method = "pearson")) %>% mutate(x="Ratio home deaths 2020 to historical mean", y = "Total cases", region = "LA"),
    broom::tidy(cor.test(home_deaths_and_cases_2020_la$ratio_annual_deaths_to_historical_mean, home_deaths_and_cases_2020_la$crude_rate_positive_per100k, method = "pearson")) %>% mutate(x="Ratio home deaths 2020 to historical mean", y = "Cases per 100k", region = "LA"),
    broom::tidy(cor.test(home_deaths_and_cases_2020_hb$ratio_annual_deaths_to_historical_mean, home_deaths_and_cases_2020_hb$cumulative_positive, method = "spearman")) %>% mutate(x="Ratio home deaths 2020 to historical mean", y = "Total cases", region = "HB"),
    broom::tidy(cor.test(home_deaths_and_cases_2020_hb$ratio_annual_deaths_to_historical_mean, home_deaths_and_cases_2020_hb$crude_rate_positive_per100k, method = "spearman")) %>% mutate(x="Ratio home deaths 2020 to historical mean", y = "Cases per 100k", region = "HB"),
    broom::tidy(cor.test(home_deaths_and_cases_2020_la$ratio_annual_deaths_to_historical_mean, home_deaths_and_cases_2020_la$cumulative_positive, method = "spearman")) %>% mutate(x="Ratio home deaths 2020 to historical mean", y = "Total cases", region = "LA"),
    broom::tidy(cor.test(home_deaths_and_cases_2020_la$ratio_annual_deaths_to_historical_mean, home_deaths_and_cases_2020_la$crude_rate_positive_per100k, method = "spearman")) %>% mutate(x="Ratio home deaths 2020 to historical mean", y = "Cases per 100k", region = "LA"),
    broom::tidy(cor.test(home_deaths_and_cases_2020_hb$deaths_all_causes, home_deaths_and_cases_2020_hb$cumulative_positive, method = "pearson")) %>% mutate(x="All home deaths", y = "Total cases", region = "HB"),
    broom::tidy(cor.test(home_deaths_and_cases_2020_hb$deaths_all_causes, home_deaths_and_cases_2020_hb$crude_rate_positive_per100k, method = "pearson")) %>% mutate(x="All home deaths", y = "Cases per 100k", region = "HB"),
    broom::tidy(cor.test(home_deaths_and_cases_2020_la$deaths_all_causes, home_deaths_and_cases_2020_la$cumulative_positive, method = "pearson")) %>% mutate(x="All home deaths", y = "Total cases", region = "LA"),
    broom::tidy(cor.test(home_deaths_and_cases_2020_la$deaths_all_causes, home_deaths_and_cases_2020_la$crude_rate_positive_per100k, method = "pearson")) %>% mutate(x="All home deaths", y = "Cases per 100k", region = "LA"),
    broom::tidy(cor.test(home_deaths_and_cases_2020_hb$deaths_all_causes, home_deaths_and_cases_2020_hb$cumulative_positive, method = "spearman")) %>% mutate(x="All home deaths", y = "Total cases", region = "HB"),
    broom::tidy(cor.test(home_deaths_and_cases_2020_hb$deaths_all_causes, home_deaths_and_cases_2020_hb$crude_rate_positive_per100k, method = "spearman")) %>% mutate(x="All home deaths", y = "Cases per 100k", region = "HB"),
    broom::tidy(cor.test(home_deaths_and_cases_2020_la$deaths_all_causes, home_deaths_and_cases_2020_la$cumulative_positive, method = "spearman")) %>% mutate(x="All home deaths", y = "Total cases", region = "LA"),
    broom::tidy(cor.test(home_deaths_and_cases_2020_la$deaths_all_causes, home_deaths_and_cases_2020_la$crude_rate_positive_per100k, method = "spearman")) %>% mutate(x="All home deaths", y = "Cases per 100k", region = "LA")
  ) %>%
  mutate(year = 2020) %>%
  relocate(region, year, x, y)

write.xlsx(x = correlations_home_deaths_and_cases, file = paste0(dir_paper, "/correlations_home_deaths_and_cases_by_region.xlsx"))
```



# Visualisations

## Common elements

```{r}
## modify theme with relevant font
theme_set(theme_minimal(base_size = 12) +
            theme(panel.grid.minor = element_blank(),
                  axis.title.y = element_text(margin = margin(0, 20, 0, 0)),  # remove gap to the left of y axis title and below the x axis title
                  axis.title.x = element_text(margin = margin(20, 0, 0, 0)),
                  text=element_text(family="Calibri")
            ))
```


## Showing historical trends as well as 2020

```{r}
## to put the arrows explaining historic range etc. only in one panel, need to construct a tibble to pass as data
# annotations_figure_place_of_death_explaining_historic_range <- tribble(
#    ~geom, ~place_of_death, ~label, ~x, ~y, ~xend, ~yend,
#   "text", "Hospital", "Historic range (min & max)", 29, 850, NA, NA,
#   "curve", "Hospital", NA, 29, 830, 32, 550,
#   "text", "Hospital", "Historic mean", 22, 800, NA, NA,
#   "curve", "Hospital", NA, 22, 780, 26, 510
# ) %>% 
#   mutate(across(.cols = c("x","xend"), .fns = ~compute_start_date_from_week_number(week_number = .x, year_number = 2020))) %>%
#   mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home")))  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)


figure_past_place_of_death <-
  weekly_deaths_current_and_past %>%
  filter(!place_of_death %in% c("All","Other") & year < 2020) %>%  # remove Other place of death & keep past years only
  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home"))) %>%  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
  # pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
               # names_to = "deaths_type", values_to = "deaths") %>% 
  ggplot(aes(x = week_number, colour = year, y = deaths_all_causes, group = year)) +
  # annotate(geom = "rect", xmin = compute_start_date_from_week_number(week_number = 1, year_number = 2021), xmax = compute_start_date_from_week_number(week_number = weeks_available_2021, year_number = 2021), ymin = -Inf, ymax = Inf, fill = "grey90", alpha = 0.2) +  # shade area - geom_rect doesn't work here
  # scale_x_date(date_labels = "%d-%m-%y", date_breaks = "8 weeks", limits = c(earliest_date, most_recent_date_available_2021)) +
  # geom_ribbon(data = merged_deaths_overall %>% filter(year==2020 & !place_of_death %in% c("All","Other")), aes(x = week_number, ymin = deaths_min, ymax = deaths_max), inherit.aes = FALSE, fill = "grey70", alpha = 0.5) +
  geom_line(aes(colour = as.factor(year))) +
  scale_colour_viridis_d() +
  # geom_line(aes(y = deaths_mean), linetype = "dashed", colour="grey20") +
  facet_grid(~place_of_death) +
  # scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
  #                     labels = c("All deaths", 
  #                                "Non Covid-19 deaths"),
  #                     guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
  #                     ) +
  theme(
    legend.position = "top",
    plot.caption = element_text(size = 10, colour = "gray60"),
    plot.margin = margin(5, 15, 5, 5),
    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
  ) +
  labs(x = "Date w/c",
       y = "Deaths (N)",
       subtitle = date_range_text,
       caption = paste0(c(death_registrations_text,"Other place of death not shown.",source_nrs_text), collapse="\n"),
       colour = "Cause of death"
       ) +
  ## annotate historic range
  # geom_text(data = annotations_figure_place_of_death_explaining_historic_range %>% filter(geom=="text"), aes(x=x, y=y, label=label),  colour = "grey20", size = 8 / (14/5)) +  # the size is tricky
  # geom_curve(data = annotations_figure_place_of_death_explaining_historic_range %>% filter(geom=="curve"), aes(x=x, y=y, xend=xend, yend=yend), colour = c("#4477aa", "grey20"), alpha = c(0.5, 1), curvature = 0.20, arrow = arrow(length = unit(0.2, "cm"))) +
  # annotate(geom = "text", x=ymd("2021-01-07"), y=30, label = "2021", hjust = 0, colour = "grey40") +
  NULL

figure1b <- 
  figure_past_place_of_death +
  labs(title=NULL)

save_output_file(filename = paste0(dir_paper,"/figure1b"), extensions = c(".pdf",".wmf",".png"), plot = figure1b, width = 28, height = 14, dpi = 300, units = "cm")
```

## Grayscale plot of weekly deaths w/ historical range for 2020

```{r}
figure1 <-
  merged_deaths_overall %>%
  filter(sex == "all" & age == "all" & year == 2020) %>%
  filter(!place_of_death %in% c("All","Other")) %>%
  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home"))) %>%  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)
  mutate(fill_text = "Historical range") %>%  # added for ease of using aes(fill) later
  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid, deaths_mean), 
               names_to = "deaths_type", values_to = "deaths") %>%
  mutate(deaths_type = factor(x = deaths_type, levels = c("deaths_all_causes", "deaths_non_covid", "deaths_mean"), labels = c("All deaths", "Non Covid-19 deaths", "Historical mean"))) %>%
  ggplot(aes(x = week_number_run_over)) +
  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max, fill = fill_text), alpha = 0.5) +
  geom_line(aes(y = deaths, linetype = deaths_type, colour = deaths_type)) +
  scale_colour_manual(values = c("black", "black", "gray30")) +
  scale_x_continuous(breaks = seq(0,60,by=10)) +
  ylim(0, NA) +  # start scale at 0 for better understanding the scale of deaths
  facet_grid(~place_of_death) +
  scale_fill_manual(values = c("grey70")) +
  theme(
    legend.position = "top",
    legend.title = element_blank(),
    legend.box.spacing = unit(0, "cm"),
    plot.caption = element_text(size = 10, colour = "gray60")
    ) +
  labs(x = "Week number",
       y = "Deaths (N)",
       caption = paste0(c(death_registrations_text, source_nrs_text), collapse="\n")
       ) +
  NULL

## figure for a bit less than half A4 page is approx 14*11cm
save_output_file(filename = paste0(dir_paper,"/figure1"), extensions = c(".pdf",".svg",".png"), plot = figure1, width = 28, height = 11, units = "cm", dpi = 300)
```


## Maps: home death increase & case counts

These are figure 2!

### Individual plots, titles removed

```{r}
figure2a_bw <-
  figure_map_ratio_home_deaths_2020_to_historical_hb_bw +
  scale_fill_scico(palette = 'grayC', direction = -1, labels = function(x) scales::percent(x, accuracy = 1)) +
  labs(title = NULL, x = NULL, y = NULL)

figure2b_bw <-
  figure_map_per_capita_covid_cases_2020_hb_bw +
  scale_fill_scico(palette = 'grayC', direction = -1, labels = function(x) format(x, big.mark=",")) +
  labs(title = NULL, x = NULL, y = NULL)

figure2c_bw <-
  figure_map_ratio_home_deaths_2020_to_historical_la_bw +
  scale_fill_scico(palette = 'grayC', direction = -1, labels = function(x) scales::percent(x, accuracy = 1)) +
  labs(title = NULL, x = NULL, y = NULL)

figure2d_bw <-
  figure_map_per_capita_covid_cases_2020_la_bw +
  scale_fill_scico(palette = 'grayC', direction = -1, labels = function(x) format(x, big.mark=",")) +
  labs(title = NULL, x = NULL, y = NULL)



save_output_file(filename = paste0(dir_paper,"/figure2a_bw"), extensions = c(".pdf",".wmf",".png"), plot = figure2a_bw, width = 12, height = 14, dpi = 300, units = "cm")
save_output_file(filename = paste0(dir_paper,"/figure2b_bw"), extensions = c(".pdf",".wmf",".png"), plot = figure2b_bw, width = 12, height = 14, dpi = 300, units = "cm")
save_output_file(filename = paste0(dir_paper,"/figure2c_bw"), extensions = c(".pdf",".wmf",".png"), plot = figure2c_bw, width = 12, height = 14, dpi = 300, units = "cm")
save_output_file(filename = paste0(dir_paper,"/figure2d_bw"), extensions = c(".pdf",".wmf",".png"), plot = figure2d_bw, width = 12, height = 14, dpi = 300, units = "cm")


```

### Combined plots

```{r}
figure2ab_bw <-
  wrap_plots(
    list(
      figure2a_bw + ggtitle("a)"), 
      figure2b_bw + ggtitle("b)")
      ), 
    nrow = 1)

figure2cd_bw <-
  wrap_plots(
    list(
      figure2c_bw + ggtitle("c)"), 
      figure2d_bw + ggtitle("d)")
      ), 
    nrow = 1)

save_output_file(filename = paste0(dir_paper,"/figure2ab_bw"), extensions = c(".pdf",".wmf",".png"), plot = figure2ab_bw, width = 20, height = 12, dpi = 300, units = "cm")
save_output_file(filename = paste0(dir_paper,"/figure2cd_bw"), extensions = c(".pdf",".wmf",".png"), plot = figure2cd_bw, width = 20, height = 12, dpi = 300, units = "cm")
```

