---
title: "Place of death trend in Scotland, 2026"
author: "Jan Savinc"
format: html
editor_options: 
  chunk_output_type: console
---

# Introduction

As of 2025-03-03, the NRS Covid-specific updates on weekly death registrations are no longer published. The new publication at https://www.nrscotland.gov.uk/publications/deaths-registered-weekly-in-scotland/ is updated weekly, and Table 7 contains the breakdown by place of death ("location").

> From 26 March 2026, we publish the tables only in the new format. The old-format tables have been archived and can be found here. A lookup table detailing the changes which were made to the way that weekly deaths statistics were published has been added to the new-format tables.

The new format tables now list a `Table4`, which contains "Weekly provisional figures on death registrations involving COVID-19 and on selected underlying causes of death by location of death in Scotland"

# Preliminaries

## Load packages

```{r}
library(tidyverse)
library(readxl)  # for reading excel files
library(curl)  # for downloading files
library(janitor)  # for cleaning datasets quickly
library(rvest)  # for scraping NRS website
library(ISOweek)  # for dealing with ISO 8601 weeks
library(ggtext)  # for fancier annotations
```

## Load functions

```{r}
source("./functions.R")
```


## Find latest file

```{r}
url_nrs_weekly_deaths <- "https://www.nrscotland.gov.uk/publications/deaths-registered-weekly-in-scotland/"

all_nrs_links <-
  read_html(url_nrs_weekly_deaths) %>%
  html_nodes("a") %>%
  html_attr("href") %>%  # find all links on the page
  tibble(links = .)

nrs_filename_pattern <- "weekly-deaths-.*\\.xlsx"

weekly_deaths_url_relative <-
  all_nrs_links %>%
  filter(str_detect(links, pattern = nrs_filename_pattern)) %>%
  filter(!str_detect(links, pattern = "monthly")) %>%  # remove monthly data
  filter(!str_detect(links, pattern = "excess")) %>%  # remove excess deaths data
  filter(!str_detect(links, pattern = "influenz")) %>%  # remove influenza data
  pull(links)

## get absolute URL
## take the first URL in the list - there's an annual file also
weekly_deaths_url_absolute <- url_absolute(weekly_deaths_url_relative[1], base = url_nrs_weekly_deaths)
```

## Download file

```{r}
date_string <- strftime(today(), format = "%F")
path_today <- file.path("./downloaded_data", date_string)
path_to_downloaded_file <- file.path(path_today, basename(weekly_deaths_url_absolute))

if (!dir.exists(path_today)) dir.create(path_today)

# download.file(url = weekly_deaths_url_absolute, destfile = path_to_downloaded_file, quiet = TRUE)
todays_file <- curl_fetch_disk(weekly_deaths_url_absolute, path = path_to_downloaded_file)
```

## Read file

```{r}
deaths_by_pod_raw <- 
  # skip first 5 rows
  read_excel(path = path_to_downloaded_file, sheet = "Table4", skip = 5) %>%
  clean_names()
```

## Set up output directory

Because data are updated on a weekly basis, we'll have a separate directory for every week the data are updated:

```{r}
latest_data_modified_date <- strftime(todays_file$modified, format="%F")
latest_year <- year(latest_data_modified_date)

dir_outputs <- file.path("./outputs", latest_year, latest_data_modified_date)
if (!dir.exists(dir_outputs)) dir.create(dir_outputs, recursive = TRUE)
```


# Wrangle data

## Convert new data to tidy (long) format

```{r}
deaths_by_pod_clean <-
  deaths_by_pod_raw %>%
  pivot_longer(cols = all_causes:covid_19_mentions, names_to = c("cause_of_death"), values_to = "n_deaths") %>%
  # recode_place_of_death() %>%
  transmute(
    year = as.integer(registration_year),
    week_number,
    date_w_c = as_date(week_beginning),
    place_of_death = location,
    year = as.integer(registration_year),
    # cause_of_death = case_when(
    #   type_of_cause == "all_causes" ~ "All",
    #   type_of_cause == "covid_19_mentioned" ~ "Covid-19 mentioned",
    #   TRUE ~ NA_character_
    # )
    cause_of_death,
    n_deaths
  )
```


## Data 2020-2025

There's a new file provided for historical context, combining 2020-2025 data for Covid, and 2000-2025 for influenza. Unfortunately only the Covid (2020-) data include data on location, so that's what I can work with.

Note: from 2026, there is a new place of death code for Hospice, which was previously partly Care home and partly Hospital. The data are labelled for whether the Hospital or Care Home is pre-2026 or 2026 and onwards.

```{r}
historical_weekly_deaths_url_absolute <- url_absolute(weekly_deaths_url_relative[2], base = url_nrs_weekly_deaths)

path_to_downloaded_historical_file <- file.path("./downloaded_data", basename(historical_weekly_deaths_url_absolute))

historical_file <- curl_fetch_disk(historical_weekly_deaths_url_absolute, path = path_to_downloaded_historical_file)

deaths_by_pod_2020_2025 <-
  # skip first 5 lines
  read_excel(path = path_to_downloaded_historical_file, sheet = "Table4", skip = 5) %>%
  clean_names %>%
  pivot_longer(cols = all_causes:covid_19_mentions, names_to = c("cause_of_death"), values_to = "n_deaths", values_transform = as.integer) %>%
  transmute(
    year = as.integer(registration_year),
    week_number,
    date_w_c = as_date(week_beginning),
    place_of_death = location,
    year = as.integer(registration_year),
    cause_of_death,
    n_deaths
  ) %>%
  recode_place_of_death()
```

## Data 2015-2019

```{r}
deaths_2015_to_2019 <- 
  read_csv("./historical_data/2020-04-30_weekly-deaths-by-location-2015-2019.csv", skip = 2) %>%
  rename(year = 1) %>%  # rename first column
  mutate(
    place_of_death = if_else(is.na(`1`), true = year, NA_character_)
  ) %>%
  fill(place_of_death, .direction="down") %>%  # LOCF on place_of_death
  drop_na(`1`) %>%
  pivot_longer(cols = all_of(as.character(1:53)), values_to = "n_deaths", names_to = "week_number") %>%
  mutate(week_number=as.integer(week_number)) %>%
  recode_place_of_death() %>%
  mutate(cause_of_death = "all_causes")
```


## Annual totals so far

```{r}
deaths_annual <-
  bind_rows(
    deaths_2015_to_2019 %>% filter(year!="average" & place_of_death!="average") %>% mutate(year=as.integer(year), cause_of_death="All"),
    deaths_current %>% filter(cause_of_death=="All")
  ) %>%
  count(year, place_of_death, cause_of_death, wt = n_deaths, name = "n_deaths") %>%
  group_by(year, place_of_death != "All") %>%
  mutate(  # add percentage
    prop_deaths = n_deaths / sum(n_deaths)
  ) %>%
  ungroup %>%
  mutate(year = if_else(  # rename current year so it's clear when the data end
    year == max(deaths_current$year),
    paste0(year, " up to W", deaths_current %>% filter(year==max(deaths_current$year)) %>% pull(week_number) %>% max),
    as.character(year)
  ))
```

## New ADR S theme

```{r}
adr_s_colours_leading <- c("#F8667D", # coral
                           "#3E0D22", # dark coral
                           "#FFFFFF"  # white
                           )

adr_s_colours_supporting <- c("#C5406E", # medium coral
                           "#32D2A7", # teal
                           "#121212",  # dark grey
                           "#F0F0F0"  # light grey
                           )
```


# Analyses

## Home deaths plot

```{r}
# TODO: separate the data wrangling into separate section
# TODO: add both the 2015-2019 historical range as a ribbon, and 2020-2025 as another ribbon, overlapping and transparent
# TODO: fix the week 52/53 anomaly at the end
# TODO: make consistent object names and variables and places of death and causes of death between years
# TODO: use the ADR-S colour scheme, make a theme for it
# TODO: label the ribbons accordingly, maybe add averages?
range_2015_2019 <-
  deaths_2015_to_2019 %>% 
  filter(year!="average") %>% 
  filter(place_of_death == "Home & other non-institution") %>% 
  mutate(
    deaths_min=min(n_deaths),
    deaths_max=max(n_deaths),
    .by=c(place_of_death,cause_of_death,week_number)
    )

range_2020_2025 <-
  deaths_by_pod_2020_2025 %>% 
  filter(cause_of_death == "all_causes") %>% 
  filter(place_of_death == "Home & other non-institution") %>% 
  mutate(
    deaths_min=min(n_deaths),
    deaths_max=max(n_deaths),
    .by=c(place_of_death,cause_of_death,week_number)
    )

week_in_2026 <- max(deaths_by_pod_clean$week_number)
all_cause_home_deaths_last <- deaths_by_pod_clean %>% filter(cause_of_death=="all_causes" & str_detect(place_of_death,"Home \\/") & week_number == week_in_2026) %>% pull(n_deaths)

(ggplot(
  data = bind_rows(deaths_by_pod_clean, deaths_by_pod_2020_2025) %>%
    mutate(year = factor(year)) %>%
    filter(str_detect(place_of_death, "Home \\/ Non-institution")) %>%
    filter(cause_of_death %in% c("all_causes")) %>%
    filter(year==2026),
  aes(group = year, x = week_number, y = n_deaths)
) +
  ## ribbon for 2015-2019 data
  geom_ribbon(data = range_2015_2019, aes(x=week_number,ymin=deaths_min,ymax=deaths_max, fill = "2015-2019"), alpha = 0.3) +
  geom_ribbon(data = range_2020_2025, aes(x=week_number,ymin=deaths_min,ymax=deaths_max, fill = "2020-2025"), alpha = 0.2) +
  geom_line(linewidth = 1, aes(colour=year)) +
  theme_minimal(base_size = 14) +
  scale_fill_manual(labels = c("2015-2019","2020-2025"), values = c(adr_s_colours_leading[1], adr_s_colours_supporting[2]), guide = guide_legend(title = "Historical range", override.aes=list(alpha=0.8))) +
  scale_colour_grey(guide = guide_legend(title=NULL)) +
  scale_x_continuous(limits = c(NA,51), n.breaks = 6) +
  # 2020-2026 range
  annotate(geom="label", x = 30, y = 450, label="Current range, 2020-2026", hjust=0, fill = adr_s_colours_supporting[2], alpha = 0.4, angle = 20) +
  # 2015-2019 range
  annotate(geom="label", x = 5, y = 220, label="Pre-pandemic range, 2015-2019", hjust=0, fill = adr_s_colours_leading[1], alpha = 0.6, angle = -20) +
  # label for 2026
  annotate(geom="text", x = week_in_2026+0.5, y = all_cause_home_deaths_last, label="2026", hjust=0, vjust=0, angle = 0) +
  ylim(c(0,NA)) +
    labs(y = "N deaths at home", 
         x = "Week number", 
         title = "Deaths at home continue at Covid-19 pandemic levels",
         # subtitle =
         #   paste("Weekly deaths at home in 2026 (solid line) are within the range seen in",
         #          "<span style='background-color: coral;'>2020-2025,</span>", "approximately 30% higher than the range in",  "<span style='background-color: #F8667D;'>2015-2019.</span>"
         #   )
         subtitle = "Weekly deaths in 2026 (solid line) are within the 2020-2025 range, elevated about 30% above the 2015-2019 range."
         ) +
  # scale_linewidth_manual(values = c(1,rep(0.5, times=5),1)) +
  theme(
    plot.title = element_textbox_simple(),
    plot.subtitle = element_textbox_simple(),
    legend.position = "none"
    ) +
  NULL
)
```

## Annual deaths

```{r}
# TODO: this is copied from 2025 file, which already wasn't working well...

# uncomment to check that the assertion about Other being ~0.5% is true
# deaths_annual %>%
#   filter(place_of_death=="Other")


(fig_proportion_deaths_by_place_annual <- 
  deaths_annual %>%
  filter(place_of_death!="All") %>%
  filter(!str_starts(year, "2025")) %>%
  filter(place_of_death != "Other") %>%
  ggplot(aes(x=year, y=prop_deaths, colour=place_of_death, shape=place_of_death)) +
  geom_point(size = rel(3)) +
  geom_line(aes(group = place_of_death)) +
  geom_text(
    data = . %>% filter(!year %in% as.character(2015:2018)) %>%
      mutate(y = if_else(place_of_death=="Care home", prop_deaths - 0.03, prop_deaths + 0.03)),
    aes(
      x = year,
      y = y,
      label = scales::label_percent(accuracy = 0.1)(prop_deaths)
    ), 
    show.legend = FALSE
  ) +
  theme_minimal() +
  theme(legend.title = element_blank(), legend.position = "top") +
  scale_colour_manual(values = viridisLite::viridis(n=4, option = "D")[1:3]) +
  # scale_colour_manual(values = RColorBrewer::brewer.pal(3, "Dark2")) +
  # scale_colour_manual(values = scales::pal_dichromat(name = "Categorical.12")(3)) +
  scale_y_continuous(labels = scales::label_percent(accuracy = 1), limits = c(0, NA)) +
  labs(x = NULL, y = "Deaths (%)", caption = '"Other institutions" represent ~0.5% of annual deaths and are not shown.'))

fig_proportion_deaths_by_place_annual %>%
save_output_file(filename = file.path(dir_outputs, paste0(c("fig_proportion_deaths_by_place_annual", latest_data_modified_date), collapse = "_")), extensions = c(".png"), plot = ., width = 20, height = 10, dpi = 300, units = "cm")
```

