---
title: "Visualisations for short paper, Feb 2023"
author: "Jan savinc"
format: html
editor_options: 
  chunk_output_type: console
---

# Load packages

```{r, warning=FALSE, message=FALSE}
## install packages from github if not yet available!
# remotes::install_github("datasciencescotland/opendatascot", force = TRUE)
# remotes::install_github("Health-SocialCare-Scotland/phsmethods", force = TRUE)

library(tidyverse)  # for tidy workflow
library(opendatascot)  # importing data from ScotGov open data website
library(phsmethods)  # methods for working with PHS data
library(readxl)  # for reading excel files
# library(SPARQL)  # taken care of by opendatascot
library(lubridate)  # dealing with dates
library(janitor)  # for cleaning column names
library(ISOweek)  # for computing date from ISO week number + year
library(sf)  # for mapping
library(ggrepel)  # for 'mapping'repelling' labels/texst in ggplot
library(patchwork)  # for assembling plots
library(extrafont)  # for working with fonts
library(openxlsx)  # for creating xlsx files
library(knitr)  # for displaying tables in rmd
library(lemon)  # for showing axis & axis labels in every facet in ggplot
library(ggtext)  # for showing coloured title text

## load functions
source("./functions.R")
```


# Copyright attribution

All data used for this study were obtained from the National Records Scotland (NRS) and are © Crown copyright, 2020; the details of the licence can be viewed on the [Open Government Licence website](http://www.nationalarchives.gov.uk/doc/open-government-licence/open-government-licence.htm)



# Dates represent event registration, not occurrence!

From NRS website:
  
  > All routine vital events information we publish is based on the date of registration, not the date on which the event occurred. Inevitably, of course, there are delays between the occurrence of an event and its registration.


# Load most recent data

```{r}
load(file = "./workspace.RData")
```


# Set up output directory

Because data are updated on a weekly basis, we'll have a separate directory for every week the data are updated:

```{r}
dir_paper <- file.path(dir_outputs, "short_paper_feb2023")
if (!dir.exists(dir_paper)) dir.create(dir_paper)
```


# Visualisations

## Common elements

```{r}
loadfonts(device="win")  # load imported fonts
# Sys.setenv(R_GSCMD = "C:/Program Files/gs/gs9.27/bin/gswin64c.exe")  # set the location of the GS file  # this not needed, just use device=device_cairo for pdfs in ggsave
windowsFonts("Calibri" = windowsFont("Calibri"))  # to suppress the multiple error messages

theme_set(theme_minimal(base_size = 12) +
            theme(panel.grid.minor = element_blank(),
                  axis.title.y = element_text(margin = margin(0, 20, 0, 0)),  # remove gap to the left of y axis title and below the x axis title
                  axis.title.x = element_text(margin = margin(20, 0, 0, 0)),
                  text=element_text(family="Calibri"),
                  axis.text = element_text(size = 9)
            ))


date_range_text <- glue::glue("Data from w/c {format(earliest_date, '%d %B %Y')} to w/c {format(most_recent_date_available_2023, '%d %B %Y')} (inclusive).")

dashed_line_historical_range_text <- "Range of deaths 2015-19 shown as shaded area; mean shown as dashed line."

only_2020_text <- glue::glue("Data from 2020 included, starting with w/c {format(earliest_date)}.")
only_2023_text <- glue::glue("Data for 2023 only include weeks 1-{format(weeks_available_2023)} (w/c {format(most_recent_date_available_2023)}).")

source_nrs_text <- "Source: National Records of Scotland"
source_phs_text <- "Source: Public Health Scotland"

crude_rate_text <- "Total cases = crude cumulative positive cases over time."

death_registrations_text <- "Note: all figures are based on date of registration."

vertical_line_text <- "Vertical dashed line at week number 54 represents start of 2021."

five_year_historical_mean <- "For 2020-2021 data, the 5-year historical period is 2015-2019; for 2022, the period is 2016-2019 + 2021, skipping 2020."

vertical_scale_differs_between_panels_text <- "Note: vertical axis not on the same scale between panels!"

# removed_other_category_text <- glue::glue("Excluding 'Other' institutional place of death, N={merged_deaths_overall %>% filter(sex == 'all' & age == 'all' & place_of_death == 'Other') %>% .$deaths_all_causes %>% sum} for weeks 1-{max(merged_deaths_overall %>% filter(year==2020) %>% .$week_number)} in 2020 and 1-{max(merged_deaths_overall %>% filter(year==2021) %>% .$week_number)} in 2021.")
removed_other_category_text <- glue::glue("Excluding 'Other' institutional place of death.")

historic_period_2015_2019_text <- "Historic mean & range for 2015-2019 throughout."
historic_period_2022_text <- "2020-21 use 2015-2019 historic mean & range; 2022 uses 2016-2019 & 2021."
```


# Place of death image

## Wrangle data

```{r}
merged_deaths_overall_2015_2019_historic_period <-
  bind_rows(
    ## pre-2022 data
    merged_deaths_overall %>% filter(year < 2022),
    ## 2022 data
    merged_deaths_overall %>%
      filter(year >= 2022) %>%  # 2022 or later!?
      select(-matches("deaths\\_mean|deaths\\_min|deaths\\_max")) %>%  # remove data using NRS historic period
      mutate(historic_period = "2015-2019") %>%  # join the previous hsitoric period data instead
      left_join(historical_weekly_average_overall, by = c("historic_period","week_number","place_of_death")) %>%
      compute_moving_average_for_current_data(  # not strictly required here but will be easier to apply in other places if included
        current_data = ., 
        historical_data = 
          weekly_deaths_2015_2019_overall %>% mutate(year=as.integer(year)) %>% drop_na(year),  # remove the entries with 'average' for year
        num_weeks_to_average = 4)
  )
```


## With 2015-2019 historic range only

Note that the footnote to the image should be separate to the file so I'll remove the title and captions and export them as a text file instead.

```{r}
fig_paper_pod_2015_2019_range <-
  merged_deaths_overall_2015_2019_historic_period %>%
  filter(sex == "all" & age == "all") %>%
  filter(!place_of_death %in% c("All","Other")) %>%
  mutate(place_of_death = factor(place_of_death, levels = order_of_place_of_death_levels[c(1,3,2)])) %>%
  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
               names_to = "deaths_type", values_to = "deaths") %>% 
  ggplot(aes(x = date_w_c)) +
  add_vertical_lines_and_annotations_for_years(years = 2021:current_year, y_annotation = 30) +
  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "12 weeks", limits = c(earliest_date, most_recent_date_available_2023)) +
  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max, fill = "grey70"), alpha = 0.5) +
  geom_line(aes(y = deaths, colour = deaths_type)) +
  geom_line(aes(y = deaths_mean, linetype = "dashed"), colour="grey20") +
  facet_grid(~place_of_death) +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths", 
                                 "Non-Covid deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  theme(
    legend.position = "top",
    legend.title = element_blank(),
    plot.margin = margin(5, 15, 5, 5),
    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
  ) +
  labs(x = NULL,
       y = "Deaths (N)"
       ) +
  ## legend to include historic range & mean
  scale_linetype_manual(values = c("dashed"), labels = "Historic weekly mean") +
  scale_fill_manual(values = c("grey50"), labels = "Historic weekly range") +
  NULL

fig_paper_pod_2015_2019_range %>%
save_output_file(filename = file.path(dir_paper, paste0(c("fig_paper_pod_2015_2019_range", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 28, height = 14, dpi = 300, units = "cm")

fig_paper_pod_2015_2019_range_footnote <- c(
  date_range_text,
  paste0(c(death_registrations_text,historic_period_2015_2019_text,source_nrs_text), collapse="\n")
)

fig_paper_pod_2015_2019_range_footnote %>%
  write_lines(file = file.path(dir_paper, paste0("fig_paper_pod_2015_2019_range_", latest_data_modified_date, ".txt")))
```


### As above, but split into 1a,b,c

```{r}
p1 <- 
  merged_deaths_overall_2015_2019_historic_period %>%
  filter(sex == "all" & age == "all") %>%
  filter(place_of_death %in% c("Home & other non-institution")) %>%
  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
               names_to = "deaths_type", values_to = "deaths") %>% 
  ggplot(aes(x = date_w_c)) +
  add_vertical_lines_and_annotations_for_years(years = 2021:current_year, y_annotation = 30) +
  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "12 weeks", limits = c(earliest_date, most_recent_date_available_2023)) +
  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max, fill = "grey70"), alpha = 0.5) +
  geom_line(aes(y = deaths, colour = deaths_type)) +
  geom_line(aes(y = deaths_mean, linetype = "dashed"), colour="grey20") +
  facet_grid(~place_of_death) +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths", 
                                 "Non-Covid deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  theme(
    legend.position = "top",
    legend.title = element_blank(),
    plot.margin = margin(5, 15, 5, 5),
    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
  ) +
  labs(x = NULL,
       y = "Deaths (N)"
       ) +
  ## legend to include historic range & mean
  scale_linetype_manual(values = c("dashed"), labels = "Historic weekly mean") +
  scale_fill_manual(values = c("grey50"), labels = "Historic weekly range") +
  NULL

p2 <-
  merged_deaths_overall_2015_2019_historic_period %>%
  filter(sex == "all" & age == "all") %>%
  filter(place_of_death %in% c("Hospital")) %>%
  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
               names_to = "deaths_type", values_to = "deaths") %>% 
  ggplot(aes(x = date_w_c)) +
  add_vertical_lines_and_annotations_for_years(years = 2021:current_year, y_annotation = 30) +
  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "12 weeks", limits = c(earliest_date, most_recent_date_available_2023)) +
  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max, fill = "grey70"), alpha = 0.5) +
  geom_line(aes(y = deaths, colour = deaths_type)) +
  geom_line(aes(y = deaths_mean, linetype = "dashed"), colour="grey20") +
  facet_grid(~place_of_death) +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths", 
                                 "Non-Covid deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  theme(
    legend.position = "top",
    legend.title = element_blank(),
    plot.margin = margin(5, 15, 5, 5),
    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
  ) +
  labs(x = NULL,
       y = NULL
       ) +
  ## legend to include historic range & mean
  scale_linetype_manual(values = c("dashed"), labels = "Historic weekly mean") +
  scale_fill_manual(values = c("grey50"), labels = "Historic weekly range") +
  NULL

p3 <-
  merged_deaths_overall_2015_2019_historic_period %>%
  filter(sex == "all" & age == "all") %>%
  filter(place_of_death %in% c("Care home")) %>%
  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
               names_to = "deaths_type", values_to = "deaths") %>% 
  ggplot(aes(x = date_w_c)) +
  add_vertical_lines_and_annotations_for_years(years = 2021:current_year, y_annotation = 30) +
  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "12 weeks", limits = c(earliest_date, most_recent_date_available_2023)) +
  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max, fill = "grey70"), alpha = 0.5) +
  geom_line(aes(y = deaths, colour = deaths_type)) +
  geom_line(aes(y = deaths_mean, linetype = "dashed"), colour="grey20") +
  facet_grid(~place_of_death) +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths", 
                                 "Non-Covid deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  theme(
    legend.position = "top",
    legend.title = element_blank(),
    plot.margin = margin(5, 15, 5, 5),
    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
  ) +
  labs(x = NULL,
       y = NULL
       ) +
  ## legend to include historic range & mean
  scale_linetype_manual(values = c("dashed"), labels = "Historic weekly mean") +
  scale_fill_manual(values = c("grey50"), labels = "Historic weekly range") +
  NULL

fig_paper_pod_2015_2019_range_1abc <-
  p1 + (p2 / p3) +
    plot_layout( 
      guides = "collect",
      widths = c(2,1)
    ) +
    plot_annotation(tag_levels = "a", tag_suffix = ")") &
    # if (interactive()) theme_minimal() else NULL &  # for testing
    theme(
      legend.position = "bottom", 
      plot.tag = element_text(size = rel(1.1))
      )

fig_paper_pod_2015_2019_range_1abc %>%
save_output_file(filename = file.path(dir_paper, paste0(c("fig_paper_pod_2015_2019_range_abc_layout", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png",".tiff"), plot = ., width = 28, height = 14, dpi = 300, units = "cm")

fig_paper_pod_2015_2019_range_1abc_footnote <- c(
  "a) Home & other non-insitution, b) Hospital, c) Care home; 'Other' place not shown due to small numbers.",
  fig_paper_pod_2015_2019_range_footnote
)

fig_paper_pod_2015_2019_range_1abc_footnote %>%
  write_lines(file = file.path(dir_paper, paste0("fig_paper_pod_2015_2019_range_1abc_footnote_", latest_data_modified_date, ".txt")))
```

# Place of death by LA

```{r}
fig_proportion_of_deaths_at_home_by_la <-
  proportions_of_deaths_by_place_by_geography %>%
  filter(geography == "la") %>%
  filter(place_of_death == "Home & other non-institution") %>%
  left_join(merged_proportions_of_deaths_by_place %>% select(year, place_of_death, proportion_scotland = proportion_of_total), by = c("year","place_of_death")) %>%
  mutate(ref_area = factor(x = ref_area, levels = order_la_by_population)) %>%  # arrange la/la by population size
  pivot_longer(cols = c(proportion_of_total, proportion_scotland), names_to = "measure", values_to = "proportion_of_total") %>%
  mutate(measure = if_else(measure == "proportion_of_total", true = "Health board", false = "Scotland")) %>%
  ggplot(., aes(x=year, y=proportion_of_total, linetype = measure)) +
  geom_point() +
  geom_line() +
  geom_text(data = ~filter(.x, year==latest_year), aes(x=latest_year+0.2, y=proportion_of_total, label="*"), inherit.aes = FALSE) +
  facet_rep_wrap(~ ref_area, repeat.tick.labels = TRUE) +
  scale_linewidth_manual(values = c(1,1.8,1,1), name = "Place of death", labels = order_of_place_of_death_levels[c(1,3,2,4)]) +
  scale_x_continuous(breaks = sort(unique(proportions_of_deaths_by_place_by_geography$year))) +
  scale_y_continuous(label = function(x) scales::percent(x, accuracy = 1)) +
  scale_colour_viridis_d(option="C", name="Place of death", labels = order_of_place_of_death_levels[c(1,3,2,4)]) +
  theme(
    legend.position = "top",
    legend.text = element_text(size = 12),
    plot.caption = element_text(size = 10, colour = "gray60"),
    panel.border = element_rect(fill = NA, colour = "gray80"),  # draw a transparent rectangle as the border
    strip.background = element_rect(fill = NA, colour = "gray80"),
    axis.text.x = element_text(angle = 45, hjust = 1)
    ) +
  labs(
    linetype = NULL,
    x = NULL, y= "Proportion of yearly deaths that happened at home or other non-institution (%)",
    subtitle = paste0("* ",only_2023_disaggregated_text),
    # caption = paste0(c("Local Authorities ordered by 2020 population estimate.",death_registrations_text,source_nrs_text),collapse="\n")
    ) +
  NULL

fig_proportion_of_deaths_at_home_by_la %>%
save_output_file(filename = file.path(dir_paper, paste0(c("fig_proportion_of_deaths_at_home_by_la", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png", ".tiff"), plot = ., width = 30, height = 25, dpi = 300, units = "cm")

fig_proportion_of_deaths_at_home_by_la_footnote <- c(
  "Local Authorities are shown in order of 2020 popoulation estimate.",
  date_range_text_disaggregated,
  paste0(c(death_registrations_text,source_nrs_text), collapse="\n")
)
fig_proportion_of_deaths_at_home_by_la_footnote %>%
  write_lines(file = file.path(dir_paper, paste0("fig_proportion_of_deaths_at_home_by_la_footnote_", latest_data_modified_date, ".txt")))
```

